---
title: "Get data from DB"
author: "SÞJ"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get data from DB}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

On survey, fish biology data are collected with the MFRI system
`HafVog` (sometimes also referred to with the english monicker
`SeaScale`) and stored in a local data base (DB) but are then
incorporated in the institute's main DB after the survey. Here we show
an example of how such data are extracted from the hafvog-scheme at
the institute on Skúlagata, but also show the minor modifications
needed to address an appropriate DB on survey. Further, since there
are more ways then one to access MFRI-DBs from R, we show two
alternative angles of attack.

## DB access

We start with the package `ROracle` (and `DBI` implicitly) and
deliberately use explicit set up of the DB-connection. Although more
low-level than using the much used (at MFRI) package `ora`, in a way a
wrapper around `ROracle`, our approach better illustrates the choice
of DB that must be made on board (when scientists are removed from the
office), and must also be changed on-shore after the survey if
analysis is to be repeated there in the aftermath of the survey.

Currently, we have to authenicate us. Here we use (and distribute) the
user and password combination of 'a databrowsing user' that is
generally set up for the the HafVog/SeaScale-DBs (and shown in the
manual for the system):

```{r, echo = TRUE, message = FALSE}
##library(ROracle)
##con <- dbConnect(dbDriver('Oracle'))
## previously it was necessary to authenticate e.g. with:
##
## user = 'hafvog_user', password = 'hafvog', dbname = 'mar')
## or:
## user = 'gagnasja', password = 'gagnasja', dbname = 'xe')
## 
## note that we would like a user/pass authentication 
## pair available on every installed HafVog and open for distribtution
## or alternatively the default user would have access to the hafvog-scheme
## dbname = 'xe') 
## on a machine with a local Oracle Extended Editoin receiving data from HafVog
## depending on setups, the user/pass combination could be skipped
## Further the hafvog manual mentions attempting:
## dbname = 'localhost:1521/xe') may be necessary
```

We collect station information in one query:

```{r, echo = TRUE, message = FALSE}
##stations <- dbGetQuery(con, "
##  select a.synis_id, stod, dags, 
##    to_char(togbyrjun,'yyyy-mm-dd HH24:MM:SS') upphaf, 
##    to_char(togendir,'yyyy-mm-dd HH24:MM:SS') endir,
##    dypi_hift, dypi_kastad, togdypi_hift, togdypi_kastad, 
##    vir_uti, larett_opnun, lodrett_opnun, grandaralengd,
##    togtimi, toglengd, toghradi,
##    kastad_n_breidd, kastad_v_lengd, hift_n_breidd, hift_v_lengd 
##  from hafvog.stodvar a, hafvog.togstodvar b
##  where a.synis_id = b.synis_id
##    and leidangur in ('A1-2016')
##  order by stod")
# change names to lowercase
##names(stations) <- casefold(names(stations))
```  

and the fish biology information in another, renaming the fields
according to the following scheme:

- a age
- l length
- w weight
- s sex
- kt maturity stage (acronymming the Icelandic term `kynthroski`)

```{r, echo = TRUE, message = FALSE}
# Kvarnir
##fishData <- dbGetQuery(con, "
##  select a.synis_id, stod, nr, lengd l, aldur a, 
##    oslaegt w, kyn s, kynthroski kt
##  from hafvog.stodvar a, hafvog.kvarnir b
##  where a.synis_id = b.synis_id 
##    and leidangur in ('A1-2016') and tegund = 31
##  order by stod, nr")
##names(fishData) <- casefold(names(fishData))
```

A bit of data manipulation is necessary, we need to:

- change to decimal degrees,

- determine split of fish into mature and immature according to
  maturity staging in field `kynthroski` (stage 1 and 2 immature,
  assume also fish were immature if maturity state is missing, all
  others mature),

- get the proportion mature at each station, and

- skip fish where age determination was missing (admittedly a bit
  sloppy, but this is usually only a fish or two).

```{r, echo = TRUE, message = FALSE}
##library(geo)
# convert to decimal degrees
##stations$lat1 <- geoconvert(stations$kastad_n_breidd)
##stations$lon1 <- -geoconvert(stations$kastad_v_lengd)
##stations$lat2 <- geoconvert(stations$hift_n_breidd)
##stations$lon2 <- -geoconvert(stations$hift_v_lengd)
## find midpoint of tow
##stations$lat <- (stations$lat1 + stations$lat2)/2
##stations$lon <- (stations$lon1 + stations$lon2)/2
# maturity categorisation
##fishData$m <- ifelse(fishData$kt %in% c(NA, 1, 2), 0, 1)
# skip those with missing age
##id <- which(is.na(fishData$a))
##fishData <- fishData[-id,]
```

Optionally, output to a file:

```{r, echo = TRUE, message = FALSE}
# saveRDS(stations, file = "../data/stations_0.rds")
# saveRDS(fishData, file = "../data/fishData_0.rds")
```
Finally, exit cleanly

```{r, echo = TRUE, message = FALSE}
##dbDisconnect(con)
```

## More ways than one to skin a cat

The github-package `fishvice/mar`, which provides functionality on top
of dplyr and dplyrOracle connection to a DB, was introduced to
haf_og_vatn/MFRI in November 2016.

Although it may not be available for all users and setups yet (e.g. on
board resarch vesselss, or collected with HafVog on commercial
vessels, port sampling from the branches, ...), we now show how this
approach can be used and how it can recuce the verbosity of our code.

Similar to before we set up a connection to the DB, in this case the
standard Skúlagata-DB (we make use of functionality built into `mar`
and utilities newly built into our Oracle DB:

```{r, echo = TRUE, message = FALSE}
##library(dplyr)
##library(dplyrOracle)
##library(mar)
##mar <- src_oracle('mar')
```

We can either query the fiskar or hafvog versions of the data, we start with the final destination of the data, it is loaded onto schema fiskar, and the mar packages 
```{r, echo = TRUE, message = FALSE}
##lesa_stodvar(mar) %>%
##  filter(leidangur == 'A1-2016') %>%
##  select(synis_id, stod, dags, 
##    togbyrjun, togendir, togtimi, toglengd, toghradi,
##    kastad_n_breidd, kastad_v_lengd,
##    hift_n_breidd, hift_v_lengd) %>%
##  mutate(lat1 = geoconvert1(kastad_n_breidd),
##    lon1 = -geoconvert1(kastad_v_lengd),
##    lat2 = geoconvert1(hift_n_breidd),
##    lon2 = -geoconvert1(hift_v_lengd)) %>%
##  collect() %>%
##  mutate(lat = (lat1 + lat2)/2,
##    lon = (lon1 + lon2)/2) -> stations2
```  

```{r, echo = TRUE, message = FALSE}
# Kvarnir
##lesa_kvarnir(mar) %>%
##  filter(tegund == 31) %>%
##  collect(n=Inf) %>%
##  right_join(stations, by = "synis_id") -> fishData2
```


## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
